Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-01-14 paper/main.tex
Updates

\section{Introduction}

Understanding cause-and-effect relationships between variables is a fundamental
goal in many scientific inquiries, as it reveals the underlying mechanisms of
observed phenomena and help in determining effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to find these relationships
among random variables from observational data. Methods for causal discovery
have been developed in both the Directed Acyclic Graphs (DAGs) and Structural
Equation Models (SEMs) literatures, each adopting a different approach.

In the DAG literature, the focus has primarily been on developing automated
algorithms to learn causal structures from datasets. As a result of these
efforts, many algorithms for causal discovery have been proposed, such as
constraint-based methods like the PC algorithm \citep{Spirtes2001} and Fast
Causal Inference (FCI) \citep{Spirtes2000}), score-based methods like
Hill-Climb Search and Greedy Equivalence Search (GES) \citep{Chickering2002},
and continuous optimization-based methods like NO TEARS \citep{Zheng2018} and
DAGMA \citep{Bello2022}. In contrast, the SEM literature has focused more
towards developing tools to assist researchers in manually constructing or
modifying models. This approach allows experts to easily integrate their expert
knowledge into the model-building process. Researchers typically develop an
initial model based on their expertise and then use various tools to guide
modifications to improve the model's fit to data. This model construction
process is commonly known as \emph{Specification Search} \citep{Long1983} and
uses method like modification indices, and Wald-based tests
\citep{Marcoulides2018}.

Despite the many automated algorithms proposed for constructing DAGs, their
adoption in applied research remains limited, with researchers often preferring
to manually construct DAGs based on their domain expertise \citep{Tennant2020,
Petersen2021}. We believe this preference persists due to several issues with
these algorithms when applied in practical settings with finite sample size:

\begin{enumerate}
	\item \textbf{Lack of Trust:} Although most algorithms are
		asymptotically consistent, their finite sample properties are
		not well understood \todo{Is there any citation for this}.
		Moreover, their output can vary significantly depending on the
		chosen algorithm and their hyperparameters. Combined with the
		lack of robust methods to test these algorithms' performance on
		a given dataset, it can be difficult to trust their outputs.
		Consequently, even when using these algorithms, users have to
		rely on post hoc manual modifications to the model based on model
		testing and their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges based on their expert knowledge before using
		them for analysis.
\end{enumerate}

@@

As a result of these issues, whether constructing the model manually or using
an algorithm, the user has to rely on their expertise to make modifications to
the model. One of the common ways to guide this post hoc modification is to use
implied CI based model testing \citep{Ankan2021, Ankan2023}. In this testing
procedure, users test whether the implied CIs of the DAG hold in the dataset or
not using statistical CI tests. Based on the results of these statistical
tests, we can accordingly make modifications to the current model to fix the
violations. However, this CI based testing also has a few drawbacks: 1) CIs are
implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example, in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

@@

\begin{figure}
	\centering
	% \includegraphics[scale=0.4]{plots/web_tool_full.png}
	\caption{A screenshot of the web tool for constructing the model. Users
		can upload their dataset after which the tool creates an empty
		graph and shows all pair of variables which are associated in
		the model using undirected red edges with the strength of
		association represented using edge width. Users can then add
		edges to the model (shown in green). Unnecessary edges are shown in 
		black.}
\end{figure}

>>>>

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to identify such relationships
among random variables using observational data. Approaches to causal discovery
have been developed within both the Directed Acyclic Graphs (DAGs) and
Structural Equation Models (SEMs) frameworks, each adapting a different approach.

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, including constraint-based methods like
the PC algorithm \citep{Spirtes2001} and Fast Causal Inference (FCI)
\citep{Spirtes2000}), score-based methods such as Hill-Climb Search and Greedy
Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. In contrast, the SEM literature has focused on developing
tools to assist researchers in manually constructing or modifying models,
enabling experts to incorporate their domain knowledge in the model building
process. Researchers typically begin with an initial model based on their
expertise and then use these tools to guide modifications that improve the
model's fit to data. This process is commonly known as \emph{Specification
Search} \citep{Long1983} and uses method such as modification indices, and
Wald-based tests \citep{Marcoulides2018}.

Despite the availability of many automated algorithms for constructing DAGs,
their adoption in applied research remains limited. Researchers often prefer to
manually construct DAGs based on their domain expertise \citep{Tennant2020,
Petersen2021}. We believe this preference persists due to several challenges
with these algorithms when applied in practical settings:

\begin{enumerate}
	\item \textbf{Lack of Trust:} While most algorithms are
		asymptotically consistent, their finite sample properties are
		not well understood \todo{Is there any citation for this}.
		Additionally, their output can vary significantly depending on
		the chosen algorithm and their hyperparameters. Combined with a
		lack of robust methods to test these algorithms' performance on
		a given dataset, it can be difficult to trust their outputs. As
		a result, even when using these algorithms are used,
		researchers often rely on post hoc manual modifications to the
		model based on model testing and their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} As multiple
		DAGs can be faithful to a given observational dataset, the
		algorithms can only recover the MECs. These MECs can contain a
		combination of directed and undirected edges. However, most
		methods for downstream tasks, such as identification or causal
		effect estimation, assume knowledge of a fully oriented DAG. As
		a result, researchers have to manually orient these edges based
		on their expert knowledge before using them for analysis.
\end{enumerate}

@@

As a result of these issues, expert knowledge integration is vital to get usable 
results from these algorithms.
% whether constructing the model manually or using
% an algorithm, the user has to rely on their expertise to make modifications to
% the model. 
One way to do this to specify expert knowledge in terms of
fixing edges, blacklisting edges, or specifying tiered knowledge.
The other ways is to do post hoc modification to the learned model using
implied CI based model testing \citep{Ankan2021, Ankan2023}. In this testing
procedure, users test whether the implied CIs of the DAG hold in the dataset or
not using statistical CI tests. Based on the results of these statistical
tests, we can accordingly make modifications to the current model to fix the
violations. However, this CI based testing also has a few drawbacks: 1) CIs are
implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example, in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

@@

\begin{figure}
	\centering
	\includegraphics[scale=0.4]{plots/web_tool_full.png}
	\caption{A screenshot of the web tool for constructing the model. Users
		can upload their dataset after which the tool creates an empty
		graph and shows all pair of variables which are associated in
		the model using undirected red edges with the strength of
		association represented using edge width. Users can then add
		edges to the model (shown in green). Unnecessary edges are shown in 
		black.}
\end{figure}
